### 2025-07-10 paper/quarto_aaai/paper.tex
MNIST image added, though stil missing IG

We find that CT generally leads to substantial and statistically
significant improvements in plausibility: average reductions in \(IP\)
range from around 7\% for \emph{MNIST} to almost 60\% for \emph{Circ};
for the real-world tabular datasets they are around 12\% for both
\emph{CH} and \emph{Cred} and almost 25\% for \emph{GMSC}; for
\emph{Adult} and \emph{OL} we find no significant impact of CT on
\(IP\). Reductions in \(IP^{*}\) are even more substantial and generally
statistically significant, although the average degree of uncertainty is
higher than for \(IP\): average reductions range from around 20\%
(\emph{Moons}) to almost 90\% (\emph{Circ}). The only negative findings
for \emph{OL} and \emph{MNIST} are statistically insignificant and, in
the latter case, do not align with qualitative findings.

@@

We also find that CT can reduce sensitivity to immutable, protected
features and thus lead to less costly counterfactual outcomes as
illustrated in Figure~\ref{fig-poc}. In column three of
Table~\ref{tbl-main}, we impose mutability constraints on selected
features and compute the reduction in average costs of counterfactuals
associated with CT compared to the weak baseline: for synthetic
datasets, we always protect the first feature; for all real-world
tabular datasets we could identify and protect an \emph{age} variable;
for \emph{MNIST}, we protect the five top and bottom rows of digits.
Reductions in costs are overwhelmingly positive and significant of up to
nearly 60\% for \emph{GMSC}. While the estimated cost reductions for
\emph{Adult} and \emph{MNIST} are not significant, FIGURE demonstrates
that CT does have the expected effect: sensitivity to protected features
as proxied by integrated gradients is drastically reduced\footnote{Additional
  findings for integrated gradients are reported in the appendix.}. In
the case of \emph{Cred}, average costs increase, likely because any
potential benefits from protecting the \emph{age} are outweighed by the
increase in costs required for greater plausibility.

>>>>

We find that CT generally leads to substantial and statistically
significant improvements in plausibility: average reductions in \(IP\)
range from around 7\% for \emph{MNIST} to almost 60\% for \emph{Circ};
for the real-world tabular datasets they are around 12\% for both
\emph{CH} and \emph{Cred} and almost 25\% for \emph{GMSC}; for
\emph{Adult} and \emph{OL} we find no significant impact of CT on
\(IP\). Reductions in \(IP^{*}\) are even more substantial and generally
statistically significant, although the average degree of uncertainty is
higher than for \(IP\): average reductions range from around 20\%
(\emph{Moons}) to almost 90\% (\emph{Circ}). The only negative findings
for \emph{OL} and \emph{MNIST} are statistically insignificant and, in
the latter case, do not align with qualitative findings, which are much
more plausible for CT (Figure~\ref{fig-mnist}).

\begin{figure}

\pandocbounded{\includegraphics[keepaspectratio]{../../paper/figures/mnist_body.png}}

\caption{\label{fig-mnist}Explanations for \emph{MNIST} for BL (top) and
CT (bottom). First columns is a randomly chosen factual 0. Columns 2
through 5 are \emph{ECCo} counterfactuals for that factual in target
classes 1 to 4. Columns 6 trough 10 show integrated gradients averaged
over all test images in the corresponding classes (5-9).}

\end{figure}%

@@

We also find that CT can reduce sensitivity to immutable, protected
features and thus lead to less costly counterfactual outcomes as
illustrated in Figure~\ref{fig-poc}. In column three of
Table~\ref{tbl-main}, we impose mutability constraints on selected
features and compute the reduction in average costs of counterfactuals
associated with CT compared to the weak baseline: for synthetic
datasets, we always protect the first feature; for all real-world
tabular datasets we could identify and protect an \emph{age} variable;
for \emph{MNIST}, we protect the five top and bottom rows of digits.
Reductions in costs are overwhelmingly positive and significant of up to
nearly 60\% for \emph{GMSC}. While the estimated cost reductions for
\emph{Adult} and \emph{MNIST} are not significant,
Figure~\ref{fig-mnist} demonstrates that CT does have the expected
effect: sensitivity to protected features as proxied by integrated
gradients is drastically reduced\footnote{Additional findings for
  integrated gradients are reported in the appendix.}. In the case of
\emph{Cred}, average costs increase, likely because any potential
benefits from protecting the \emph{age} are outweighed by the increase
in costs required for greater plausibility.

### 2025-07-10 paper/sections/experiments.qmd
MNIST image added, though stil missing IG

We find that CT generally leads to substantial and statistically significant improvements in plausibility: average reductions in $IP$ range from around 7% for *MNIST* to almost 60% for *Circ*; for the real-world tabular datasets they are around 12% for both *CH* and *Cred* and almost 25% for *GMSC*; for *Adult* and *OL* we find no significant impact of CT on $IP$. Reductions in $IP^{*}$ are even more substantial and generally statistically significant, although the average degree of uncertainty is higher than for $IP$: average reductions range from around 20% (*Moons*) to almost 90% (*Circ*). The only negative findings for *OL* and *MNIST* are statistically insignificant and, in the latter case, do not align with qualitative findings.

<!-- TODO: Add figure for MNIST -->

@@

We also find that CT can reduce sensitivity to immutable, protected features and thus lead to less costly counterfactual outcomes as illustrated in @fig-poc. In column three of @tbl-main, we impose mutability constraints on selected features and compute the reduction in average costs of counterfactuals associated with CT compared to the weak baseline: for synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five top and bottom rows of digits. Reductions in costs are overwhelmingly positive and significant of up to nearly 60% for *GMSC*. While the estimated cost reductions for *Adult* and *MNIST* are not significant, FIGURE demonstrates that CT does have the expected effect: sensitivity to protected features as proxied by integrated gradients is drastically reduced^[Additional findings for integrated gradients are reported in the appendix.]. In the case of *Cred*, average costs increase, likely because any potential benefits from protecting the *age* are outweighed by the increase in costs required for greater plausibility.

<!-- TODO: Add figure for MNIST -->

>>>>

We find that CT generally leads to substantial and statistically significant improvements in plausibility: average reductions in $IP$ range from around 7% for *MNIST* to almost 60% for *Circ*; for the real-world tabular datasets they are around 12% for both *CH* and *Cred* and almost 25% for *GMSC*; for *Adult* and *OL* we find no significant impact of CT on $IP$. Reductions in $IP^{*}$ are even more substantial and generally statistically significant, although the average degree of uncertainty is higher than for $IP$: average reductions range from around 20% (*Moons*) to almost 90% (*Circ*). The only negative findings for *OL* and *MNIST* are statistically insignificant and, in the latter case, do not align with qualitative findings, which are much more plausible for CT (@fig-mnist).

![Explanations for *MNIST* for BL (top) and CT (bottom). First columns is a randomly chosen factual 0. Columns 2 through 5 are *ECCo* counterfactuals for that factual in target classes 1 to 4. Columns 6 trough 10 show integrated gradients averaged over all test images in the corresponding classes (5-9).](/paper/figures/mnist_body.png){#fig-mnist}

@@

We also find that CT can reduce sensitivity to immutable, protected features and thus lead to less costly counterfactual outcomes as illustrated in @fig-poc. In column three of @tbl-main, we impose mutability constraints on selected features and compute the reduction in average costs of counterfactuals associated with CT compared to the weak baseline: for synthetic datasets, we always protect the first feature; for all real-world tabular datasets we could identify and protect an *age* variable; for *MNIST*, we protect the five top and bottom rows of digits. Reductions in costs are overwhelmingly positive and significant of up to nearly 60% for *GMSC*. While the estimated cost reductions for *Adult* and *MNIST* are not significant, @fig-mnist demonstrates that CT does have the expected effect: sensitivity to protected features as proxied by integrated gradients is drastically reduced^[Additional findings for integrated gradients are reported in the appendix.]. In the case of *Cred*, average costs increase, likely because any potential benefits from protecting the *age* are outweighed by the increase in costs required for greater plausibility.
